Antisense transcripts can quickly evolve to encode proteins

### 2023-03-24 Manuscripts/M2/antisense_genes.tex
modified:   Manuscripts/Figures/M2_main/Divergence_scer.svg 	modified:   Manuscripts/M2/antisense_genes.tex 	modified:   Modeling/antisenseGenes.jl

%frame 1 vs intergenic 
%
%no sel:
% 30
% 30
% 33
% 46
%
%ws: 
% 43
% 54
% 70
% 93
% 
%ss:
%  49
%  63
%  81
% 109

%genomic:
%ns, ws, ss : 30, 60, 69

\section*{Discussion}

@@

>>>>

\subsection*{Overlap with sense ORFs promotes divergence of antisense ORFs in frame 2}

In the previous sections, we showed that purifying selection on the sense ORF can affect the emergence and loss of antisense ORFs. We next asked if this purifying selection can also constrain the diversification of the proteins encoded by antisense ORF sequences. To this end, we first calculated the chemical distance ($\delta$) between any two amino acids. For this calculation we used a distance matrix that we derived from an experimentally estimated amino acid similarity matrix reported in a previous study \citep{PMBEC}. Next, we calculated the average chemical difference ($\bar{\delta}$) introduced by a random mutation, weighted by the probability of different mutations. Using $\bar{\delta}$ as a measure of divergence, we estimated the extent to which antisense ORFs in the three frames, can diverge as a result of mutations, and due to purifying selection on the sense ORFs. Likewise, we also calculated the divergence of intergenic ORFs as a consequence of random mutations.

We found that frame 2 allows maximum divergence of antisense ORFs, under both weak and strong purifying selection on the sense ORF. Antisense ORFs in frame 0 diverge the least. Interestingly, strong selection on sense ORF increases the divergence of antisense ORFs in frame 2. The reason could be that the few mutations that do occur under strong purifying selection, cause a relatively higher increase in divergence than the more numerous mutations that are allowed to occur under weak purifying selection. We also found that the divergence of frame 2 antisense ORFs was higher than than that of intergenic ORFs under both selection regimes. We note this result does not mean that intergenic ORFs can diverge less than antisense ORFs. Evolution of intergenic ORFs is not constrained by another DNA sequence. However, as long as the mutants do not affect the organismal fitness, evolution would not be biased towards divergence increasing mutations. Thus random mutations in intergenic ORFs could also consist of many synonymous and chemistry preserving mutations, that are probably disallowed in frame 2 antisense ORFs due to purifying selection on sense ORFs.

In contrast to frame 2, the divergence of antisense ORFs in the other two frames decreased with increasing strength of purifying selection on the sense ORF. For example, frame 0 antisense ORFs did not diverge at all when the sense ORF was under strong purifying selection. Antisense ORFs in frames 0 and 1 also diverged less than intergenic ORFs under both selection regimes.

\begin{figure}[!t]
\centering
\includegraphics[scale=0.5]{\figdir/Divergence_scer.pdf}
\caption{Antisense ORFs can diversify when sense ORFs are under purifying selection. Vertical axis denotes the divergence ($\bar{\delta}$) of antisense ORFs due to a random mutation when the sense ORF is under weak (left) or strong (right) purifying selection. Horizontal axes denote the three antisense frames. Colored bars denote divergence values of antisense ORFs with different GC-content, and black bars denote the diversity values calculated using frequencies of short DNA sequences from the yeast genome. Filled circles that are similarly color coded, denote the divergence of intergenic ORFs due to mutations.}
\label{divergence}
\end{figure}
\section*{Discussion}

@@

\subsection{Divergence of protein sequence}
Specifically, if $i$ denotes the original codon, $j$ denotes the substituted codon, $P_i$ denotes the probability of finding codon-$i$, $\mu_{ij}$ denotes the probability of codon-$i$ mutating to codon-$j$, and $\delta_{ij}$ denotes the chemical difference between the amino acids encoded by these codons, then the average chemical difference is defined by the following equation:

\begin{equation}
\bar{\delta} = \frac{\displaystyle\sum_i P_i \sum_j \mu_{ij}\delta_{ij}}{\displaystyle\sum_i P_i \sum_j \mu_{ij}}
\end{equation}
